File-based localization of numerical perturbations in data analysis pipelines

### 2019-12-16 paper.tex

\title{Spot: a tool to 
locate numerical differences in pipelines}

\begin{document}

\author[1]{Ali Salari}
\author[1]{Lalet Scaria}
\author[2,3]{Gregory Kiar}
\author[2]{Lindsay Lewis}
\author[2,3]{Alan C. Evans}
\author[1]{Tristan Glatard}

@@

Many experiments show that computational analysis results are still not 
completely reproducible. Computational environments including different 
operating systems, hardware categories, and software versions are known 
to have effects on the results produced by analysis pipelines. These 
effects are presumably due to the creation, propagation and 
amplification of small numerical differences across the pipelines. 
Although new techniques such as virtualization, version control 
systems, and provenance management tools provide a more reliable 
environment and significantly improve the reproducibility of scientific 
findings. However, the precise causes of such instabilities and the 
path along which they propagate in the pipelines are unclear.  We 
present a technique to identify the processes in the pipeline that 
create numerical differences along with the execution, and we apply this 
technique to the HCP structural pre-processing pipelines.

@@

\begin{keywords}
Reproducibility; Numerical Instability; Neuroimaging.
\end{keywords}

@@

Reproducibility is a crucial element of the scientific works, 
as it enables researchers to evaluate authenticity and reliability 
of the findings~\cite{plesser2018reproducibility}.
Reproducibility is defined as the ability to regenerate the same 
results as the original findings when the experiment is reanalyzed by 
the same analytic methods, software package, parameters, and 
data~\cite{peng2011reproducible}. 
In addition, numerical reproducibility is defined as the ability to 
regenerate bit for bit identical results from multiple 
runs~\cite{hill2017numerical}. 
We considered the numerical reproducibility in our experiments by comparing 
the binary content of the results using the checksum method.

Recently, validation of the reproducibility has been widely investigated 
in the field of neuroimaging in which uses optimized 
processing methods with the purpose of functional and structural 
assessments of the human brain.
According to the previous 
studies, the variety of computing infrastructures including workstation 
types, parallelization methods, operating systems, and analysis 
packages are known to influence the reproducibility of the analyses~\cite{Gronenschild2012, 
diethelm2012limits, Glatard2015, bowring2019exploring}.
These irreproducibility issues are reported as the result of the 
creation, propagation, and amplification of small numerical 
differences.

In particular, the effect of the operating systems on  
computational pipelines shows the creation of small numerical differences~\cite{Glatard2015, Scaria2017}.
These differences mainly correspond to the mathematical functions implemented 
in different operating system libraries.
For instance, changing the mathematical functions like \emph{expf()} and 
\emph{cosf()} which manipulate the precision of floating-point representations, 
between \emph{glibc} libraries in different operating systems can produce 
small numerical differences.
Besides, a similar issue is expected for any operating system which is 
based on \emph{glibc}, the GNU C library.

There are different approaches to improve the reproducibility of the analysis, 
including containerization techniques that encapsulate software/hardware dependencies, 
provenance capturing tools, and version control systems. 
However, a comprehensive solution requires to fix the numerical instabilities instead of 
masking the problem. For this purpose, bagging is one solution that has been used to 
stabilize motion estimation analysis in fMRI~\cite{Glatard2018hbm}. 
Bootstrap aggregation reduces the instabilities, but it needs expensive computations.
As the second solution, a debugging tool can help to identify and 
fix the instabilities in the pipeline. Therefore, we introduce 
the numerical reproducibility measurement tool named \emph{Spot} to identify 
the processes in the pipeline that create numerical differences across different conditions. 
Furthermore, we can stabilize the identified processes in the next steps.

The remainder of this paper describes the Spot tool in different parts 
including the way of capturing and representation of the provenance 
information, the method of clustering of different subject types, and 
process labelling to characterize differences. We describe the HCP 
pre-processing pipelines as a popular neuroimaging project to test the 
Spot tool and evaluate its functionality. Finally, we reported the 
experimental results which describe the processes that are responsible 
for the differences in the pipelines.

@@

@@

>>>>

\title{File-based localization of numerical perturbations in data analysis pipelines}
\begin{document}

\author[1]{Ali Salari}
\author[2,3]{Gregory Kiar}
\author[2]{Lindsay Lewis}
\author[2,3]{Alan C. Evans}
\author[1]{Tristan Glatard}

@@

Operating systems and other computational factors are known to have an
effect on the results produced by neuroimaging pipelines, presumably due to
the creation, propagation and amplification of numerical errors along data
analysis pipelines. However, the precise causes of such instabilities and
the path along which they propagate in the pipelines are unclear, while
this process could play a major role in the current reproducibility crisis.
We present a technique to identify the computational processes in a
pipeline that create numerical differences when executed in different
computational conditions. Our technique leverages system-call interception to reconstruct
processing and file graphs without pipeline instrumentation. Applying our tool to
the structural pre-processing pipelines released by the Human Connectome
Project, we find that linear and non-linear registration are the cause of
most numerical instabilities in such pipelines, which confirms previous
findings.

@@

\begin{keywords}
Computational reproducibility; Numerical instability; Neuroimaging.
\end{keywords}

@@

% Reproducibility: crisis, numerical stability
% Containerization isn't the solution

Numerical stability is believed to play a central role in the differences
observed in data analysis pipelines across computational infrastructures.
Variability resulting from operating system upgrades, parallelization, or
even minor data perturbations may originate in errors that accumulate in
numerical computations, with important consequences on the reproducibility
of data analysis results. However, it remains challenging to identify the
exact elements in a complex pipeline that are prone to numerical instability,
leaving pipeline developers and users with the daunting task to manually
compare results obtained in different conditions.

In neuroimaging, our primary application field, data analyses often consist
of hundreds of computational processes, sometimes coming from multiple
toolboxes, that are aggregated to perform a specific function. For
instance, the fMRIprep pipeline~\cite{esteban2019fmriprep} assembles software blocks
from FSL, AFNI, Freesurfer and ANTs to provide a state-of-the art
functional MRI processing tool with minimal user input. Another example are the pipelines of the Human
Connectome Project~\cite{glasser2013} that combine tools from FSL and Freesurfer to pre-process
structural, functional and diffusion data from their
open dataset. In both cases, the pipelines leverage toolboxes that are
widely trusted in the community. At the same time, substantial results
variations have been observed in these toolboxes resulting from minor data
or infrastructure perturbations~\cite{Glatard2015, Gronenschild2012, Lewis2017-ll}, which suggests that further investigation of their
numerical conditioning is required. For such complex pipelines, a
lightweight solution has to be found to perform such evaluations with
limited code instrumentation.

Numerical evaluations are traditionnally performed using techniques such as
interval arithmetics~\cite{hickey2001interval} that require complete code re-writes and are
therefore barely applicable to complex pipelines. Recently, Monte-Carlo
Arithmetic~\cite{Parker1997-qq, Denis2016-wo} provided a practical way to
evaluate the uncertainty of numerical results without the need to rewrite
the application in a different paradigm. By perturbating floating-point
computations, it introduces a controllable amount of noise in the
pipelines, effectively sampling results from a random distribution. While
this technique is very appealing, it suffers from two main issues that make
it impractical at the scale of a complete pipeline. First, it requires that
all software components be recompiled for MCA instrumentation, which is not
always feasible. Second, it multiplies the execution time by a factor of 10
to 100, which is impractical when executions already take a few hours to
complete.

We present Spot, a tool to identify the source of numerical differences in
complex pipelines without instrumentation. Using system-call interception
through the Reprozip tool~\cite{chirigati2016reprozip}, Spot traverses graphs of processes and
intermediary files to pinpoint the pipeline components that are unstable
across execution conditions. When differences start accumulating,
effectively masking any further instability, it restores clean data copies
through a set of wrapper scripts. Wrapper scripts are also used to restore
temporary data that might have been deleted during the execution, and to
disambiguate files that have been written by multiple processes. The remainder of this paper 
presents the design of our tool, and its 
application to pre-processing pipelines of the HCP project.

% % Verificarlo and MCA, require recompilation
% Numerical instabilities originate in the limited \todo{precision} of floating-point representations. 
% Traditionnaly, it has been evaluated through

% % -> File-based analysis of pipelines without instrumentation
% % Reprozip

% Reproducibility is a crucial element of the scientific works, 
% as it enables researchers to evaluate authenticity and reliability 
% of the findings~\cite{plesser2018reproducibility}.
% Reproducibility is defined as the ability to regenerate the same 
% results as the original findings when the experiment is reanalyzed by 
% the same analytic methods, software package, parameters, and 
% data~\cite{peng2011reproducible}. 
% In addition, numerical reproducibility is defined as the ability to 
% regenerate bit for bit identical results from multiple 
% runs~\cite{hill2017numerical}. 
% We considered the numerical reproducibility in our experiments by comparing 
% the binary content of the results using the checksum method.

% Recently, validation of the reproducibility has been widely investigated 
% in the field of neuroimaging in which uses optimized 
% processing methods with the purpose of functional and structural 
% assessments of the human brain.
% According to the previous 
% studies, the variety of computing infrastructures including workstation 
% types, parallelization methods, operating systems, and analysis 
% packages are known to influence the reproducibility of the analyses~\cite{Gronenschild2012, 
% diethelm2012limits, Glatard2015, bowring2019exploring}.
% These irreproducibility issues are reported as the result of the 
% creation, propagation, and amplification of small numerical 
% differences.

% In particular, the effect of the operating systems on  
% computational pipelines shows the creation of small numerical differences~\cite{Glatard2015, Scaria2017}.
% These differences mainly correspond to the mathematical functions implemented 
% in different operating system libraries.
% For instance, changing the mathematical functions like \emph{expf()} and 
% \emph{cosf()} which manipulate the precision of floating-point representations, 
% between \emph{glibc} libraries in different operating systems can produce 
% small numerical differences.
% Besides, a similar issue is expected for any operating system which is 
% based on \emph{glibc}, the GNU C library.

% There are different approaches to improve the reproducibility of the analysis, 
% including containerization techniques that encapsulate software/hardware dependencies, 
% provenance capturing tools, and version control systems. 
% However, a comprehensive solution requires to fix the numerical instabilities instead of 
% masking the problem. For this purpose, bagging is one solution that has been used to 
% stabilize motion estimation analysis in fMRI~\cite{Glatard2018hbm}. 
% Bootstrap aggregation reduces the instabilities, but it needs expensive computations.
% As the second solution, a debugging tool can help to identify and 
% fix the instabilities in the pipeline. Therefore, we introduce 
% the numerical reproducibility measurement tool named \emph{Spot} to identify 
% the processes in the pipeline that create numerical differences across different conditions. 
% Furthermore, we can stabilize the identified processes in the next steps.

@@

% mention fuzzy libmath as a way to do this across distributions.

@@

% importance of data-driven evaluations, subject clustering, etc

% future work with MCA

% link registration to our motion estimation study
